Generating Random Logic Programs Using Constraint Programming

### 2020-02-11 paper/paper.tex
Rearranging stuff

Most current approaches to generating random logic programs are restrictive,
e.g., limited to clauses with only two literals
\citep{DBLP:conf/lpnmr/NamasivayamT09}, or to clauses of the form $a \gets \neg
b$ \citep{DBLP:journals/tocl/WenWSL16}, but others are more expressive, e.g.,
defining a program only by the (maximum) number of atoms in the body and the
total number of rules \citep{DBLP:conf/iclp/ZhaoL03}. We introduce a way to
generate random logic programs using a constraint solver such as Choco
\citep{choco}. The same model can generate both probabilistic programs directly
in the syntax of ProbLog \citep{DBLP:conf/ijcai/RaedtKT07} as well as
non-probabilistic Prolog programs. For generated probabilistic programs to be
valid, we use a custom constraint to detect negative cycles (as described in
\cref{sec:cycles}). A major advantage of our constraint-based approach is that
one can easily add additional constraints to the model. To demonstrate that, in
\cref{sec:independence} we present a custom constraint with propagation and
entailment algorithms that can ensure predicate independence.
\cref{sec:counting} also presents a combinatorial argument for correctness,
where we produce combinatorial expressions that count the number of programs
that the model should produce for various parameter values. Finally,
\cref{sec:experiments} shows how the model scales when tasked with producing
more complicated programs and identifies the relationships between parameter
values and the empirical hardness of the program generation task.
 % TODO (Fazl): % Could you provide a motivating example? It % might be useful to engage readers.

@@

\begin{constraint}
Let
  \[
    P = \{ h.\variable{predicate} \mid h \in \variable{heads} \}.
  \]
  Then
  \[
    \variable{nValues}(P) =
    \begin{cases}
      |\predicates{}| + 1 & \text{if } \variable{count}(\Box, P) > 0 \\
      |\predicates{}| & \text{otherwise.}
    \end{cases}
  \]
\end{constraint}
% TODO (Fazl): explain what h is.

@@

% TODO (Fazl): the first four words sound odd. Rephrase.
It remains to constrain the forest represented by the $\variable{structure}$
array together with its $\variable{values}$ to eliminate unnecessary symmetries
and adhere to our desired format. First, we can recognise that the order of the
elements in the $\variable{structure}$ array does not matter, i.e., the
structure is only defined by how the elements link to each other, so we can add
a constraint saying that:

@@

In order to demonstrate the correctness of the model and explain it in more
detail, in this section we are going to derive combinatorial expressions for
counting the number of programs with up to $\maxNumClauses{}$ clauses and up to
$\maxNumNodes{}$ nodes per clause, and arbitrary $\predicates{}$,
$\arities{}$, $\variables{}$, and $\constants{}$. To simplify the task, we only
consider clauses without probabilities and disable (negative) cycle elimination.
It was experimentally confirmed that the model agrees with the combinatorial
formula from this section in 985 different scenarios. The \emph{total arity} of
a body of a clause is the sum total of arities of all predicates in the body.
% TODO (Vaishak, last sentence): move to footnote and add "Since the [details] of
% this empirical investigation are not crucial to the thrust of this paper, we
% will [detail] this in an extendedd technical report."

@@

% TODO: talk about randomness. The value heuristic is random because we want
% randomness. After finding one solution, one can continue the search for more,
% but they will be similar (acceptable if the search space is small and we want
% to find all solutions).
% On the other hand, with restarts, duplicates become possible.

@@

Another issue worth discussing is that of randomness. Most random models sample
from a well-defined probability distribution whereas a constraint model, while
potentially sufficiently random, provides no such guarantees. In future work, it
would be interesting to examine the probability distribution of logic programs
as generated by our model, although it is not obvious how such a
probability distribution could be characterised.% One example of a clear bias
%exhibited by our model is that of complexity over simplicity, i.e., if a tree
%that represents the body of a clause can have up to $\maxNumNodes{}$
%nodes---because the number of possible trees with $n$ nodes grows with $n$---it
%is much more likely to have $\maxNumNodes{}$ or $\maxNumNodes{} - 1$ nodes than,
%e.g., one or two.

% TODO: We can do the same thing for predicates within the body of a clause that
% we do with variables, i.e., predicates within a clause are permutable (except
% for the head predicate), so we can have a set of 'allowed' predicates.

@@

\renewcommand{\bibsection}{\subsubsection*{References}}
\bibliography{paper}
% TODO: logic programs don't have disjunctions. Is there an easy way to
% eliminate them?

>>>>

Most current approaches to generating random logic programs are restrictive,
e.g., limited to clauses with only two literals
\citep{DBLP:conf/lpnmr/NamasivayamT09}, or to clauses of the form $a \gets \neg
b$ \citep{DBLP:journals/tocl/WenWSL16}, but others are more expressive, e.g.,
defining a program only by the (maximum) number of atoms in the body and the
total number of rules \citep{DBLP:conf/iclp/ZhaoL03}. We introduce a way to
generate random logic programs using a constraint solver such as Choco
\citep{choco}. The same model can generate both probabilistic programs directly
in the syntax of ProbLog \citep{DBLP:conf/ijcai/RaedtKT07} as well as
non-probabilistic Prolog programs. For generated probabilistic programs to be
valid, we use a custom constraint to detect negative cycles (as described in
\cref{sec:cycles}). A major advantage of our constraint-based approach is that
one can easily add additional constraints to the model. To demonstrate that, in
\cref{sec:independence} we present a custom constraint with propagation and
entailment algorithms that can ensure predicate independence.
\cref{sec:counting} also presents a combinatorial argument for correctness,
where we produce combinatorial expressions that count the number of programs
that the model should produce for various parameter values. Finally,
\cref{sec:experiments} shows how the model scales when tasked with producing
more complicated programs and identifies the relationships between parameter
values and the empirical hardness of the program generation task.

@@

\begin{constraint}
Let
  \[
    P = \{ h.\variable{predicate} \mid h \in \variable{heads} \}.
  \]
  Then
  \[
    \variable{nValues}(P) =
    \begin{cases}
      |\predicates{}| + 1 & \text{if } \variable{count}(\Box, P) > 0 \\
      |\predicates{}| & \text{otherwise.}
    \end{cases}
  \]
\end{constraint}

@@

It remains to constrain the forest represented by the $\variable{structure}$
array together with its $\variable{values}$ to eliminate unnecessary symmetries
and adhere to our desired format. First, we can recognise that the order of the
elements in the $\variable{structure}$ array does not matter, i.e., the
structure is only defined by how the elements link to each other, so we can add
a constraint saying that:

@@

In order to demonstrate the correctness of the model and explain it in more
detail, in this section we are going to derive combinatorial expressions for
counting the number of programs with up to $\maxNumClauses{}$ clauses and up to
$\maxNumNodes{}$ nodes per clause, and arbitrary $\predicates{}$,
$\arities{}$, $\variables{}$, and $\constants{}$\footnote{We checked that the
  our model agrees with the derived combinatorial formula in close to a thousand
  different scenarios. The details of this empirical investigation are omitted
  as they are not crucial to the thrust of this paper.}. To simplify the task,
we only consider clauses without probabilities and disable (negative) cycle
elimination. We also introduce the term \emph{total arity} of a body of a clause
to refer to the sum total of arities of all predicates in the body.

@@

% TODO: Extra notes.
% 1. Randomness. The value heuristic is random because we want randomness. After
% finding one solution, one can continue the search for more, but they will be
% similar (acceptable if the search space is small and we want to find all
% solutions).
% 2. Logic programs don't have disjunctions and parentheses. Is there an easy
% way to eliminate them?
% 3. We can do the same thing for predicates within the body of a clause that we
% do with variables, i.e., predicates within a clause are permutable (except for
% the head predicate), so we can have a set of 'allowed' predicates.
% 4. One example of a clear bias exhibited by our model is that of complexity
% over simplicity, i.e., if a tree that represents the body of a clause can have
% up to $\maxNumNodes{}$ nodes---because the number of possible trees with $n$
% nodes grows with $n$---it is much more likely to have $\maxNumNodes{}$ or
% $\maxNumNodes{} - 1$ nodes than, e.g., one or two.

@@

Another issue worth discussing is that of randomness. Most random models sample
from a well-defined probability distribution whereas a constraint model, while
potentially sufficiently random, provides no such guarantees. In future work, it
would be interesting to examine the probability distribution of logic programs
as generated by our model, although it is not obvious how such a
probability distribution could be characterised.

@@

\renewcommand{\bibsection}{\subsubsection*{References}}
\bibliography{paper}
